A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-20 jacobian_calc_CMAME.tex
beginning to address I.5

\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}
In order to illustrate the complex-step technique in a simpler form than it
appears in the \emph{Peridigm} code written for this study, an example
header-only library and two example problem definition programs that use it
have been created. All the materials described here are available in the
associated project repository.

@@

Where $\phi$ is a constant, beam angle, $AE$ is rigidity, $D$ is displacement,
$L_{0}$ is original beam length, and $k_{s}$ is the Hooke's constant for the
coil spring attached to the beam. For a verification of each of the methods as
well as a convergence rate study, this truss and spring model was analyzed in
the second program. $f(D)$ is restorative force to balance downward force
loading at the connection point of the beam and the spring, parallel to the
wall. As in the reference, equilibrium solutions for 12 incremental load steps
of 4.448 Newtons were sought. Differing from the reference, Newton's method was
used rather than 'Dynamic Relaxation'. Solver parameters common to all tests
were a limit of 100 iterations for convergence, a residual change tolerance of
$1\times{10^{-9}} \si{\pico\meter}$ and an update multiplier of 1.0. Data
collected included number of iterations to converge summed over all load-steps
and final displacement, D, after the final load-step. Additionally
finite-difference step-size was varied in order to measure the impact on each
of the Jacobian calculation methods. Naturally, varying step size does not
affect AD or the analytically derived Jacobian. The accuracy results for
verifying the Jacobian calculation methods are summarized in Table
\ref{tab:ConvergenceStudy}.

@@

Table \ref{tab:ConvergenceStudy} indicates that when using AD or the 'actual'
analytically derived Jacobian, a minimum number of iterations is achieved
corresponding to a maximum convergence rate. CS and FD are similarly affected
by a grossly large step-size, while subtractive cancellation invalidates the
results for FD for very small step-sizes. Inaccuracy for large step-sizes is is
predicted by the FD and CS formulas, which are contingent upon a small
step-size to allow the neglect of higher order 12 terms. The results of this
test show that step-size selection is not unimportant when using FD, but when
properly configured any of the methods may return the same maximum convergence
rate for this problem.  Overall conclusions regarding the methods that may be
kept in mind for problems besides this example are that the accuracy of the
Jacobian should not be confused with the accuracy of the equi- librium
solution, since each of the methods discussed here may be equally capable given
enough iterations. Additionally, while the accuracy of the Jacobian may
strongly affect convergence rates, this relationship was not quantified here.

@@

\section{Conclusions}
%
Based on the experimental results, it appears that for users of \emph{Peridigm}
there is a case to be made for using AD for reasons of performance and
accuracy. However, certain parallel tests done in this study showed a small
performance gain over AD with the FD method.  While the results showed that CS
produced more accurate tangent-stiffness matrices than CD and FD under the
parameters of the tests, it was not determined whether or not this is a clear
advantage of CS over FD in terms of accuracy of final predicted displacements
and speed of convergence. The reader is reminded that the time required to
complete an iteration must be considered along with the total number of
iterations taken order to measure speed of convergence. The goal of the
comparative study was to benchmark single iteration performance and not
convergence rate. For new developments it is recommended that combined accuracy
and speed of iteration measurements be conducted along with entirely separate
convergence studies. It should be mentioned that a related work,
\cite{perez2012numerical}, compares the speed and accuracy of complex-step to
finite difference for producing tangent-stiffness matrices in solid mechanics
simulations using a complex inelastic material model, where the researchers
similarly found CS to be accurate yet relatively computationally expensive
compared to FD and CD. In that work, full convergence studies were also done
and indicated that quadratic convergence could be achieved with complex-step
\remove[Michael]{all of the methods} \cite[p.28]{perez2012numerical}.

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated infomation stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
ammount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibilty of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation.

>>>>

\note[Michael]{re: I.5, this is taken from the thesis.}
\add[Michael]{
\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}
In order to illustrate the complex-step technique in a simpler form than it
appears in the \emph{Peridigm} code written for this study, an example
header-only library and two example problem definition programs that use it
have been created. All the materials described here are available in the
associated project repository.

@@

Where $\phi$ is a constant, beam angle, $AE$ is rigidity, $D$ is displacement,
$L_{0}$ is original beam length, and $k_{s}$ is the Hooke's constant for the
coil spring attached to the beam. For a verification of each of the methods as
well as a convergence rate study, this truss and spring model was analyzed in
the second program. $f(D)$ is restorative force to balance downward force
loading at the connection point of the beam and the spring, parallel to the
wall. As in the reference, equilibrium solutions for 12 incremental load steps
of 4.448 Newtons were sought. Differing from the reference, Newton's method was
used rather than 'Dynamic Relaxation'. Solver parameters common to all tests
were a limit of 100 iterations for convergence, a residual change tolerance of
$1\times{10^{-9}} \si{\pico\meter}$ and an update multiplier of 1.0. Data
collected included number of iterations to converge summed over all load-steps
and final displacement, D, after the final load-step. Additionally
finite-difference step-size was varied in order to measure the impact on each
of the Jacobian calculation methods. Naturally, varying step size does not
affect AD or the analytically derived Jacobian since those are not functions
of step size. The accuracy results for
verifying the Jacobian calculation methods are summarized in Table
\ref{tab:ConvergenceStudy}.

@@

Table \ref{tab:ConvergenceStudy} indicates that when using AD or the 'actual'
analytically derived Jacobian, a minimum number of iterations is achieved
corresponding to a maximum convergence rate. CS and FD are similarly affected
by a grossly large step-size, while subtractive cancellation invalidates the
results for FD for very small step-sizes. Inaccuracy for large step-sizes is is
predicted by the FD and CS formulas, which require a small
step-size to allow the neglect of higher order 12 terms. The results of this
test show that step-size selection is not unimportant when using FD, but when
properly configured any of the methods may return the same maximum convergence
rate for this problem.  Overall conclusions regarding the methods that the accuracy of the
Jacobian should not be confused with the accuracy of the equilibrium
solution, since each of the methods discussed here may be equally capable given
enough iterations. Numerical measurements of accuracy besides number of iterations taken
are not possible here because there is not basis for comparison, as the methods are solving
the problem separately and producing difference Newton iterates. Additionally, while the inaccuracy of the Jacobian may
strongly affect convergence rates, this relationship was not precisely quantified here.
}

@@

\note[Michael]{Re: I.5}
\add[Michael]{
\section{Convergence and displacement accuracy studies} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of
the methods, it was necessary to conduct convergence rate studies in order to
get a complete picture. It is necessary for an analyist to know if the
accuracy differences measured in the per-iteration tests would amount to
differences in convergence rates among the methods if they were allowed to solve problems
normally. The question is posed, given that step-size $h$ is heuristically
selected by \emph{Perdigm} for the finite difference methods and presumable
makes a good choice, do the single-iteration Jacobian accuracy results predict
the ranking of the methods in terms of convergence rate for a given study? For
brevity, convergence rate was measured for each of the methods in solving just
the largest problem of the single-core test series. As was seen in the per-iteration results,
Jacobian accuracy did not vary greatly from test to test, such that any pair of tests
from the two series could have been considered reasonably representative. As in
the validation problem shown in ~\ref{subsec:Validation}, would each of the methods return an
equal convergence rate so long as $h$ were optimally selected?

The results shown in Table ~\ref{tab:ConvergenceStudy2} indicate that this was the case. 
The difference in Jacobian accuracy between the methods seen in the per-iteration tests
was not sufficient to lead to differences in the convergence rate of the methods.

\begin{table*}[!ht]   
    \centering \caption{Average iterations/ load-step} 
    \label{tab:ConvergenceStudy2}   
    \begin{tabular}{c c c c c}
    \toprule Test type & AD & CS & CD & FD\\
        \midrule 
        single-core & 63 & " & " & "\\ 
        multi-core  & 63 & " & " & "\\ 
\bottomrule \end{tabular} \end{table*}

The end displacement results shown in ~\ref{subsec:Validation} indicated that up to a certain precision the
solutions produced by each of the methods were identical. It was examined if this was the case for the 
solutions produced in \emph{Peridigm}. Here each of the solutions produced by each of the 
methods were compared to the AD produced solutions, under the assumption that the AD produced solutions
would be closest to the solution produced by using an 'exact' Jacobian. Table ~\ref{tab:PeridigmSolutionAccuracy} shows that the methods
did indeed produce differing solutions in terms of nodal force density, although only in decimal places at least $9$ orders 
of magnitude smaller than the force density value at the node where the difference was detected.

\begin{table*}[!ht]   
    \centering \caption{Nodal force density maximum difference} 
    \label{tab:PeridigmSolutionAccuracy}   
    \begin{tabular}{c c c c}
    \toprule  & CS & CD & FD\\
        \midrule  Order of magnitude & $\num{1.E-8}$ & $\num{1.E-8}$ & $\num{1.E-5}$ \\ 
\bottomrule \end{tabular} \end{table*}

The solution accuracy results do follow the same ranking seen in the Jacobian accuracy results, however,
again as in the validation problem the results differ only slightly. Speculatively this degree of error
may be important to those running probabilistic simulations such that the propagation of error may be 
a concern. Also it is conceivable that if this error is a fixed quantity, insensitive to material 
properties or loading conditions, it may have a greater impact on simulations involving very low
force densities such as for small displacements or very lax materials. In this case it could be recommended
that forward-difference be avoided.

\section{Conclusions}
%
\subsection{For further studies of Jacobian accuracy in Newton's method}
\label{subsec:FurtherStudies}
The question is raised; what is the significance of Jacobian estimation
accuracy to the convergence of Newton's method?  The per-iteration and even
convergence studies shown here indicate the effect of Jacobian accuracy for
only a single datum, a very accurate Jacobian estimate, while the validation
problem indicates the effect of using a grossly inaccurate Jacobian. The results here show that
for very small Jacobian estimate inaccuracies, convergence rate is not
affected, while for grossly inaccurate Jacobians Newton's method fails. 
Is it worthwhile then to investigate the effect of using moderately inaccurate Jacobians, 
produced deliberately or otherwise, on the convergence rate of Newton's method?

The answer may be no at least unless a more highly non-linear class of problem
is selected for study or a controllable method of producing a moderately
inaccurate Jacobian is found along with a reason for using it. While it is possible
to incorrectly apply finite-differencing and CS by choosing an inappropriate
step-size, there is no reason that this should be done. The experiments show
that step-size $h$ can be properly selected and that the methods are capable of
producing accurate Jacobians. It should be noted that the model used to perform
the simulation could always be substituted for a more highly non-linear one and
perhaps this would somewhat increase the importance of Jacobian accuracy. The
study presented here failed to pose the question of how wrong a correctly
implemented Jacobian calculation scheme could go for plastic or otherwise
non-linear material models, here instead non-linearity was imposed by the
nonlocal interaction between material points essential to peridynamic models,
like the linear model used in the study.

Nevertheless, Newton's method is not defined for using an inaccurate or 'wrong'
Jacobian, so using an inaccurate Jacobian is failing to perform Newton's method
by definition. But it is also true that it should not be expected that every
differentiation technique performs flawlessly under all situations. The
pragmatic strategy to study this may be to compare a prospective method's
Jacobian accuracy to that produced by AD and develop a rule of thumb based on
observed convergence rates from prior simulations as they relate to the percent
difference of the prospective and AD produced Jacobians. Also it is possible to
calculate a sensitivity of Jacobian error to step-size alongside a sensitivity
of convergence rate to step-size, and then to relate these two quantities as a
map of Jacobian error to convergence rate change from ideal. Additionally it is
still possible to look at the convergence conditions for Newton's method and
suggest a relationship between Jacobian accuracy and convergence rate in a
theoretical manner. To this point, there are a set of convergence conditions derived in
\cite[Chap. 2, p.41]{burkeLectures} which relate quality of Jacobian inverse approximations
to convergence rate classification, here reproduced,
%
\begin{equation} \mid \mid (r \prime (x_{k})^{-1} - J_{k}^{-1})y_{k} \mid \mid \leq K,
\end{equation}

for worse than linear convergence,

\begin{equation} \mid \mid (r \prime (x_{k})^{-1} - J_{k}^{-1})y_{k} \mid \mid \leq \Theta_{1}^{k}K; \Theta_{1} \in (0,1),
\end{equation}

for linear convergence,

\begin{equation} \mid \mid (r \prime (x_{k})^{-1} - J_{k}^{-1})y_{k} \mid \mid \leq min{M_{2} \mid \mid x_{k} - x_{k-1} \mid \mid, K}; M_{2} > 0
\end{equation}

for two step quadratic,

\begin{equation} \mid \mid (r \prime (x_{k})^{-1} - J_{k}^{-1})y_{k} \mid \mid \leq min{M_{3} \mid \mid r(x_{k}) \mid \mid, K}; M_{3} > 0
\end{equation}

and lastly for quadratic convergence. Here $r \prime$ is the true system
Jacobian and $J$ is the approximation. Essentially the conditions measure how
well the difference of the true Jacobian inverse and approximation works as
a contraction map within the closure of a radius of convergence about the true
solution. This property is necessary for the solution method using the
approximation to converge rather than to diverge, as is true of the exact Newton
method.

It may be interesting to artificially perturb the Jacobian calculations with
random error in order to better control experimentally the onset of inaccuracy
to develop a sensitivity of convergence rate to this type of error. Perhaps an
analyst could use this knowledge to devise an adaptive differentiation
technique selection method which economizes on accuracy until a significant
Jacobian error is measured during a special doubly calculated iteration.

Additionally, there are Quasi-Newton methods such as Broyden's method which
uses an approximate inverse Jacobian and updates it with information from the 
iterate and residual \cite{dennis1971convergence}. A method called
Incomplete Jacobian Newton's method may the closest to the idea of deliberately
making use of an inaccurate Jacobian in the sense that arbitrary elements of
the full Jacobian are set to zero to reduce computational effort
\cite{liu2008incomplete}. The Quasi-Newton methods represent a compromise of
Newton's method for the promise of faster iterations in exchange for a lower
convergence rate. For both of these methods, the iterate and residual are
computed as in Newton's method. As to the question of algorithmic consistency,
these numerical methods evaluate the same linearized constitutive law and
governing equation as for the exact Newton's method which in the case of a
plastic simulation are involved in iteratively determining algorithmic tangent moduli. 
In fact, a secant equation based method is appropriate for Newton's method with algorithmic
tangent moduli since the Jacobian ought only depend on known converged
values such that the Newton iterate could not artificially drive the plasticity
of the model \cite[Chap. 6]{belytschko1999nonlinear}. It should be noted that
constraints or a line search algorithm to minimize the residual are necessary
whereas the pure secant method does not return unique solutions in
multi-dimensions.
%

\subsection{Grain of salt for speed results}
The per-iteration speed results shown here indicate that the AD method is
faster than CD or FD.  While this may be the case for the implementation in the
stock version of \emph{Peridigm} this is not reflective of any limitation of
these methods. In a side experiment, the finite-difference code in
\emph{Peridigm} was re-written to take advantage of the benefits of storing the
results of duplicated intermediate calculations. These modifications were 
general, low memory footprint and could be adopted in any numerical simulation
simulation software as good practice. An informal performance test was done
using a personal AMD Phenom-2 X4 workstation using GCC 4.7 compilers rather than a
Intel Xeon E5 node with Intel Cluster Studio 13.0 compilers as provided by TACC. The problem
being run was the largest single core test from the main study, instead run on four 
compute cores. Table ~\ref{tab:PerformanceMod} shows the timing
results for AD as reference along with an unmodified CD, and then modified CD
and FD.

\begin{table*}[!ht]   
\centering \caption{Total time spent computing Jacobian} 
\label{tab:PerformanceMod}   
\begin{tabular}{c c c c c}
\toprule & AD & CD & Modded CD & Modded FD\\
\midrule  Total seconds & $~430$ & $~570$ & $~430$ &$~320$ \\ 
\bottomrule \end{tabular} \end{table*}

The results indicated that in exchange for a modest programming effort, a
performance increase of around a quarter was available for the
finite-difference methods and that FD could eclipse AD. What allowed the performance modifications to
enhance the finite-difference methods was their lack of sophistication
compared to AD in that their explicitly laid out and
repetitive code could be readily optimized.

\subsection{Remarks}
Based on the experimental results, it appears that for users of \emph{Peridigm}
there is a case to be made for using AD for reasons of speed. However, certain
parallel tests done in this study showed a small performance gain over AD with
the FD method.  The results showed that CS produced more accurate
tangent-stiffness matrices than CD and FD under the parameters of the tests,
however it was determined that this difference in accuracy was not sufficient
to alter convergence rate for a test problem. In order for Jacobian accuracy to
have an impact on convergence rate, and therefor serve as a predictor of
convergence rate in per-iteration results, the amount of error detected between
a supposed perfectly accurate Jacobian and the test Jacobian needs to be
greater than allowed by the conditions seen in \cite{burkeLectures}, 
reproduced in \ref{sec:FurtherStudies}.

It should be mentioned that another work, \cite{perez2012numerical}, compares
the speed and accuracy of complex-step to finite difference for producing
tangent-stiffness matrices in solid mechanics simulations using the 'MRS-Lade'
material model, where the researchers similarly found CS to be accurate yet
relatively computationally expensive compared to FD and CD. In that work, full
convergence studies were also done and indicated that quadratic convergence was
achieved with the methods \cite[p.28]{perez2012numerical}.

However, at least for the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated. 
}
The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated infomation stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
ammount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibilty of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation. \add[Michael]{However, at least for 
the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated.}
